---
title: "Single-cell ATAC-seq, Session 4"
subtitle: " <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> Bioinformatics Resource Center - Rockefeller University"
author:
  - "[http://rockefelleruniversity.github.io/scATAC/](http://rockefelleruniversity.github.io/scATAC/)"
  - "[brc@rockefeller.edu](mailto:brc@rockefeller.edu)"
---

```{r setup, include=FALSE, purl=FALSE}
suppressPackageStartupMessages(require(knitr))
knitr::opts_chunk$set(echo = TRUE, tidy = T)
```

::: {.content-hidden when-format="revealjs"}
# Single-cell ATAC-seq: Differential Accessibility and Visualization
:::

## Overview

- [Recap](#recap)
- [Differential accessibility](#differential-accessibility)
- [Pseudobulk differential accessibility](#pseudobulk-differential-accessibility)
- [Visualizing regions of interest](#visualizing-regions-of-interest)

## Recap

?? Brief recap of Session 3 — LSI, clustering, gene activity scores and label transfer ??

- [Session 3](Session3.html)

# Differential accessibility {background-color="#23373B" .center .middle}

## Differential accessibility

?? What question differential accessibility answers, and the choice between cell-level and sample-level (pseudobulk) testing ??

## FindMarkers

?? Cell-level testing between clusters/cell types; choice of test (e.g. LR with fragment-count latent variable) and its tendency to inflate significance by treating cells as replicates ??

```{r findmarkers, eval=FALSE}
## ?? e.g. FindMarkers(test.use = "LR", latent.vars = "nCount_peaks") ??
```

## Annotating differential peaks

?? Linking differential peaks to nearby genes ??

```{r closest_feature, eval=FALSE}
## ?? e.g. Signac::ClosestFeature() ??
```

# Pseudobulk differential accessibility {background-color="#23373B" .center .middle}

## Pseudobulk differential accessibility

?? Aggregating counts per sample × cell type so biological replicates are the unit of testing; requires multiple samples per condition ??

```{r pseudobulk, eval=FALSE}
## ?? e.g. AggregateExpression() then DESeq2 ??
```

## Comparing approaches

?? Where FindMarkers and pseudobulk results agree and disagree; report adjusted p-values ??

# Visualizing regions of interest {background-color="#23373B" .center .middle}

## Coverage plots

?? Visualizing pseudobulk accessibility tracks per group over a region, with gene annotation and peaks ??

```{r coverage_plot, eval=FALSE}
## ?? e.g. Signac::CoveragePlot(region = ..., extend.upstream = ..., extend.downstream = ...) ??
```

## Customizing coverage plots

?? Adding peak, link and expression panels; combining tracks ??

## Time for an exercise!

Exercise on differential accessibility and visualization can be found [here](../../exercises/exercises/Session4_Exercise_exercise.html)

## Answers to exercise.

Answers can be found [here](../../exercises/answers/Session4_Exercise_answers.html)
