---
title: "Single-cell ATAC-seq, Session 3"
subtitle: " <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> Bioinformatics Resource Center - Rockefeller University"
author:
  - "[http://rockefelleruniversity.github.io/scATAC/](http://rockefelleruniversity.github.io/scATAC/)"
  - "[brc@rockefeller.edu](mailto:brc@rockefeller.edu)"
---

```{r setup, include=FALSE, purl=FALSE}
suppressPackageStartupMessages(require(knitr))
knitr::opts_chunk$set(echo = TRUE, tidy = T)
```

::: {.content-hidden when-format="revealjs"}
# Single-cell ATAC-seq: Dimensionality Reduction and Clustering
:::

## Recap

Thus far we have covered:
- Cell Ranger 
- Importing data into R and Signac
- Generating QC metrics
- Deciding on cutoffs
- Normalization and feature selection

## Overview

- [Latent Semantic Indexing](#latent-semantic-indexing)
- [Embedding and clustering](#embedding-and-clustering)
- [Revisiting QC](#revisiting-qc)
- [Gene activity scores](#gene-activity-scores)
- [Label transfer from scRNA-seq](#label-transfer-from-scrna-seq)


# Latent Semantic Indexing {background-color="#23373B" .center .middle}

## Latent Semantic Indexing

LSI = TF-IDF followed by singular value decomposition (SVD). 

Just as a reminder this helps deal with some of the specifics of scATAC data structure. Thus far we have done TF-IDF which normalizes for per-cell sequencing depth (TF) and up-weights rarer, more informative peaks (IDF).

The next step is SVD. This in place of PCA for scRNAseq.

## Why Can't We Use PCA?

In scRNA-seq, **PCA** is the standard dimensionality reduction step applied after log-normalization.

For scATAC-seq, the TF-IDF matrix is:

- **Non-negative** — values range from 0 upwards
- **Sparse** — most entries are 0
- **High-dimensional** — up to ~100,000 peaks

PCA applied directly to TF-IDF data performs poorly because PCA assumes approximately Gaussian-distributed inputs and is sensitive to the distribution of zeros.

## What is SVD doing?

SVD summarises ~100,000 peaks as a few dozen **"accessibility programs"**:
groups of peaks that tend to open and close together across cells
(e.g. a set of T-cell enhancers). The top dimensions capture the major axes of chromatin accessibility variation across cells — analogous to principal components.

It splits the TF-IDF matrix into three parts:

| Part | Plain meaning |
|---|---|
| **U** | Which peaks make up each program |
| **Σ** | How much variation each program explains |
| **V** | How much of each program each cell has |

We keep **V**: each cell becomes ~50 numbers (one per program) instead of
~100,000 peaks. These are the **LSI components** we use for UMAP and clustering.

For more details on [SVD](https://www.pnas.org/doi/10.1073/pnas.97.18.10101), there is a good explainer for when it was first applied to genomics for microarrays. 

## Running SVD in Signac

Lets start by loading in the object as we left it (or continue from where you were in the session).

```{r runSVD}
pbmc <- readRDS("data/pbmc_filtered_tfidf.rds")

pbmc <- RunSVD(pbmc)

pbmc[["lsi"]]

```

## SVD dimensions

`RunSVD()` defaults to 50 dimensions. 

* The **first singular vector** in LSI typically is almost always strongly correlated with total fragment count (sequencing depth) rather than meaningful biological variation. This is because cells with high reads tend to have fewer non-zero entries across all peaks.
* You rarely need all 50 for clustering. 

Dimensions 2–30 are a common working set, but we can always check `DepthCor()` first.

```{r depth_cor, eval=FALSE}
DepthCor(pbmc)
```

## Inspecting DepthCor Output

The output is a bar plot showing the Pearson correlation of each LSI component with `log10(nCount_peaks)`.

```{r depthCor}
DepthCor(pbmc)
```

::: {.callout-important}
## What to look for
- **LSI 1:** correlation close to +/-1 → **always exclude**
- **Later components:** correlations > 0.75 → may also be depth-driven; consider excluding
- **Most datasets:** skip LSI 1, use 2–30

Downstream calls that use LSI must specify `dims = 2:N` explicitly.
:::


# UMAP Embedding {background-color="#23373B" .center .middle}

## From LSI to UMAP

::: {.columns}

::: {.column width="50%"}
**UMAP** projects the high-dimensional LSI space into 2D for visualisation. Compared to t-SNE it generally:

- Preserves more **global structure** (relative position of clusters matters)
- Runs faster on large datasets
- Is more reproducible with a fixed seed

We set `seed.use = 42` for reproducibility.
:::

::: {.column width="50%"}
```{r runUMAP}
pbmc <- RunUMAP(
  object    = pbmc,
  reduction = "lsi",
  dims      = 2:30,
  seed.use  = 42
)

DimPlot(object = pbmc, label = TRUE) + NoLegend()
```
:::

:::

## Interpreting the UMAP

Cells cluster together based on **similarity of their chromatin accessibility profiles**.

Different cell types have distinct combinations of accessible regulatory elements and should separate in UMAP space.


## Common UMAP pitfalls: depth gradients

Run this check before trusting the UMAP:
```{r}
pbmc2 <- RunUMAP(
  object    = pbmc,
  reduction = "lsi",
  dims      = 1:30,
  seed.use  = 42
)

FeaturePlot(pbmc2, features = "nCount_peaks", max.cutoff = "q95")
```

If you see a smooth gradient of fragment count across the UMAP, LSI component 1 may have been included. Re-run `RunUMAP()` excluding it.

# Graph-Based Clustering {background-color="#23373B" .center .middle}

## Clustering with Seurat

The clustering workflow mirrors scRNA-seq:

1. **Build a k-nearest neighbour (KNN) graph** in LSI space
2. **Refine with SNN** (shared nearest neighbour)
3. **Community detection** — Louvain, SLM or Leiden algorithm

Leiden is considered the gold standard these days, but you also need to install the [leiden algorithm outside of R](https://leidenalg.readthedocs.io/en/stable/intro.html). In the Signac vignette they use SLM, so lets go for that.

## Clustering with Seurat

```{r findClusters}
pbmc <- FindNeighbors(
  object    = pbmc,
  reduction = "lsi",
  dims      = 2:30
)

pbmc <- FindClusters(
  object    = pbmc,
  verbose   = FALSE,
  algorithm = 3,
  random.seed = 42
)
```

## Clustering with Seurat

```{r}
DimPlot(object = pbmc, label = TRUE) + NoLegend()
```

## Tuning the Resolution Parameter

There is no single "correct" resolution. The right level of granularity depends on your biological question — coarser for broad lineages, finer for rare subtypes. Try several values and compare cluster stability.

```{r resolutionTuning, eval=FALSE}
pbmc <- FindClusters(pbmc, resolution = 0.5)   # coarser
pbmc <- FindClusters(pbmc, resolution = 2.0)   # finer

p1 <- DimPlot(pbmc, group.by = "peaks_snn_res.0.5", label = TRUE) +
      ggtitle("Resolution 0.5") + NoLegend()
p2 <- DimPlot(pbmc, group.by = "peaks_snn_res.2",   label = TRUE) +
      ggtitle("Resolution 2.0") + NoLegend()
```

## Tuning the Resolution Parameter

```{r}

p1 + p2
```


# QC Revisited {background-color="#23373B" .center .middle}

## QC is Iterative, Not a Single Step

It is good practice to **re-examine QC metrics in the context of your clusters** after embedding.

Low-quality cells may form isolated clusters or cluster with real cell types but with distinctly worse metrics.

```{r qcOnUMAP1}
FeaturePlot(
  object     = pbmc,
  features   = c("nCount_peaks",
    "TSS.enrichment",
    "excl_prop",
    "nucleosome_signal"),
  ncol       = 2,
  max.cutoff = "q95",
  pt.size = 0.2
)
```

## QC is Iterative, Not a Single Step

```{r qcOnUMAP2}
FeaturePlot(
  object     = pbmc,
  features   = c(
    "pct_reads_in_peaks",
    "amulet.q",
    "scDblFinder.score"),
  ncol       = 2,
  max.cutoff = "q95",
  pt.size = 0.2
)
```

## QC is Iterative, Not a Single Step

Violin plots split by cluster can also help discern differences. 

```{r qcOnVlnPlot}
VlnPlot(
  object     = pbmc,
  features   = c("nCount_peaks",
    "TSS.enrichment",
    "excl_prop",
    "nucleosome_signal",
    "pct_reads_in_peaks",
    "scDblFinder.score",
    "amulet.q"),
  split.by = "peaks_snn_res.2",
  ncol       = 3
)
```

## Revisiting doublets and other artefacts

**Doublets** often appear between major clusters in the UMAP.

Signs of a doublet cluster:

- Intermediate accessibility profile
- Higher fragment count than neighboring clusters
- Enriched for markers of two different cell types

## Cyclical QC strategy

Let's remember: 
1. Apply conservative thresholds initially
2. UMAP and Cluster
3. Double-check QC metrics on UMAP to see if anything is sticking out
4. Update filters or remove suspect clusters. Go back to 2, until there are no remaining QC concerns. 

Two to three iterations is common before settling on a final cell set.

## Our data?

We can see some clear separation of some QC metrics across the UMAP. 

There are low counts in one cluster. But the other metrics look pretty okay. 

QC does not have to end here though. We will keep an eye on that cluster as we do check for marker peaks and do cell type annotation. 

# Gene activity scores {background-color="#23373B" .center .middle}

## From Peaks to Genes

We have information about our clusters relating to their QC, but assigning cell type labels will also be informative. Clear cell type identities will indicate pure population. To be able to do this we use a proxy for gene expression: `GeneActivity()`.

**Idea:** sum the ATAC-seq fragment counts overlapping a window around each gene's body and promoter.


::: {.callout-note}
This is a **proxy** for expression, not true RNA. Accessible chromatin at a gene promoter or body correlates with transcription but does not equal it — especially for genes regulated primarily by distal enhancers.Typically framgents that overlap 2kb upstream of TSS and whole gene body are counted. 
:::

## Computing Gene Activity

We generate gene activity scores and add them to our Signac object as a separate assay. We will largely treat this like RNA data. 

This step will recount in the context of genes, instead of peaks. This means we will need the fragments file for this.

```{r geneActivity}
gene.activities <- GeneActivity(pbmc)

pbmc[["RNA"]] <- CreateAssayObject(counts = gene.activities)

pbmc <- NormalizeData(
  object             = pbmc,
  assay              = "RNA",
  normalization.method = "LogNormalize",
  scale.factor       = median(pbmc$nCount_RNA)
)

DefaultAssay(pbmc) <- "RNA"
```

## Using Gene Activity for Annotation

```{r geneActivityViz}
FeaturePlot(
  object     = pbmc,
  features   = c(
    "MS4A1",   # B cells
    "CD3D",    # T cells
    "LEF1",    # naive T cells
    "NKG7",    # NK cells
    "TREM1",   # monocytes
    "LYZ"      # monocytes/DCs
  ),
  pt.size    = 0.1,
  max.cutoff = "q95",
  ncol       = 3
)
```

## Strengths and Limitations of Gene Activity

## Strengths
- Quick, no additional data required
- Works for well-studied cell types with known marker genes
- Provides biologically interpretable axes for initial annotation
- Scales to large datasets (computed from the fragment file directly)

## Limitations
- Not true gene expression — accessibility ≠ transcription
- Long genes with many peaks, or gene dense regions may be systematically over-represented
- Poorly captures enhancer-regulated genes with distal regulatory elements
- Noisy for rare or lowly expressed genes

## Best practice
Use gene activity as a **guide** and confirm with label transfer from a matched scRNA-seq reference where available. The two approaches are complementary.


# Label Transfer from scRNA-seq {background-color="#23373B" .center .middle}

## Leveraging a Reference scRNA-seq Dataset

There already exist many well documented scRNAseq atlases or well curated datasets for most experimental contexts. We can use our Gene Activity along with Seurats label transfer methodology to be able to use published data to label cells.

## Leveraging a Reference scRNA-seq Dataset

Seurat's **label transfer** workflow:

1. Uses an annotated scRNA-seq dataset as a **reference**
2. Maps your scATAC-seq cells onto the reference via shared gene-activity features
3. Transfers cell type labels plus a per-cell **prediction score** (confidence)

More robust than gene activity alone because it uses the full transcriptomic reference to anchor the mapping.

## Finding Transfer Anchors

```{r transferAnchors, eval=FALSE}
DefaultAssay(pbmc) <- "RNA"

pbmc_rna <- readRDS("data/pbmc_10k_rna.rds")

transfer.anchors <- FindTransferAnchors(
  reference       = pbmc_rna,
  query           = pbmc,
  features        = VariableFeatures(object = pbmc_rna),
  reference.assay = "RNA",
  query.assay     = "RNA",
  reduction       = "cca"   # canonical correlation analysis
)
```

::: {.callout-note}
CCA identifies cells that are similar in the shared gene space — these become the "anchors" used to transfer information. The gene activity matrix (stored as the "RNA" assay) serves as the bridge between the ATAC query and the RNA reference.
:::

## Transferring Labels

```{r transferLabels, eval=FALSE}
celltype.predictions <- TransferData(
  anchorset       = transfer.anchors,
  refdata         = pbmc_rna$celltype,
  weight.reduction = pbmc[["lsi"]],
  dims            = 2:30
)

pbmc <- AddMetaData(pbmc, metadata = celltype.predictions)

DimPlot(
  object   = pbmc,
  group.by = "predicted.id",
  label    = TRUE,
  repel    = TRUE
) + NoLegend()
```

## Evaluating Label Transfer Quality

```{r evalTransfer, eval=FALSE}
hist(pbmc$prediction.score.max,
     breaks = 50,
     xlab   = "Max prediction score",
     main   = "Label transfer confidence")

pbmc$low_confidence <- pbmc$prediction.score.max < 0.5
table(pbmc$low_confidence)
```

::: {.callout-tip}
## Interpreting low-confidence predictions
Cells with prediction score < 0.5 may represent:
- Cell types **absent from the reference** (→ may need a broader reference)
- **Low-quality cells** that survived QC (→ consider removing)
- **Rare transitional states** between annotated types (→ biologically interesting)

Cross-reference low-confidence cells with QC metrics before deciding whether to keep or remove them.
:::





# Label transfer from scRNA-seq {background-color="#23373B" .center .middle}

## Label transfer from scRNA-seq

?? Using an annotated scRNA-seq reference to label scATAC cells via gene activity scores and CCA anchors ??

```{r label_transfer, eval=FALSE}
## ?? e.g. FindTransferAnchors(reduction = "cca") then TransferData(weight.reduction = atac[["lsi"]]) ??
```

## Assessing transferred labels

?? Prediction scores, agreement with clusters, and when to distrust a transferred label ??

## Time for an exercise!

Exercise on dimensionality reduction and clustering can be found [here](../../exercises/exercises/Session3_Exercise_exercise.html)

## Answers to exercise.

Answers can be found [here](../../exercises/answers/Session3_Exercise_answers.html)
